---
title: "Population and Landscape Change in the Denver Metro Area"
format: html
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidycensus)
library(tidyverse)
library(tigris)
library(sf)
library(tmap)
library(scales)

census_api_key("9383d03114a6fd61b90f6aa740342df389ec984f",
               install = TRUE, overwrite = TRUE)
options(tigris_use_cache = TRUE)

# # ============================================================
# REPLICATION STUDY — Denver Metro Area
# Author: Linda Adade
# Date: 05/08/2026
# Hanberry (2022) Landscape Classification
# Walker (2023) Analyzing US Census Data
# ============================================================

# LIBRARIES
library(tidycensus)
library(tidyverse)
library(tigris)
library(sf)
library(tmap)
library(scales)

census_api_key("9383d03114a6fd61b90f6aa740342df389ec984f",
               install   = TRUE,
               overwrite = TRUE)

options(tigris_use_cache = TRUE)

# ============================================================
# STEP 1 — Download 2020 Census tracts for Colorado
# Walker Ch. 2.1.1 — get_decennial(), sumfile = "pl" for 2020
# ============================================================
co_tracts <- get_decennial(
  geography = "tract",
  variables = "P1_001N",
  state     = "CO",
  geometry  = TRUE,
  year      = 2020,
  sumfile   = "pl"
)

# ============================================================
# STEP 2 — Download Denver MSA boundary
# Walker Ch. 7.1.2 — core_based_statistical_areas()
# ============================================================
denver_msa <- core_based_statistical_areas(cb = TRUE, year = 2020) %>%
  filter(str_detect(NAME, "Denver-Aurora")) %>%
  st_transform(26954)

# ============================================================
# STEP 3 — Spatially filter tracts to Denver MSA
# Walker Ch. 7.1.2 — st_filter() with st_within predicate
# CRS 26954 = NAD83 Colorado Central (meters) for area calc
# ============================================================
co_tracts     <- co_tracts %>% st_transform(26954)
denver_tracts <- co_tracts %>%
  st_filter(denver_msa, .predicate = st_within)

# ============================================================
# STEP 4 — Calculate population density and classify landscapes
# Walker Ch. 3.3.2 — mutate() and case_when()
# Hanberry (2022) Table 4 Worldpop thresholds (p.8)
# Units: persons per square kilometer
# ============================================================
denver_density <- denver_tracts %>%
  rename(pop_total = value) %>%
  mutate(
    area_sqmi   = as.numeric(st_area(geometry)) / 2589988,
    area_km2    = area_sqmi / 0.386,
    pop_density = pop_total / area_km2,
    landscape   = case_when(
      is.na(pop_density)  ~ "Exurban",
      pop_density >  1900 ~ "Urban High-Density",
      pop_density >=  800 ~ "Urban Low-Density",
      pop_density >=  550 ~ "Suburban High-Density",
      pop_density >=  250 ~ "Suburban Low-Density",
      TRUE                ~ "Exurban"
    ),
    landscape = factor(landscape, levels = c(
      "Urban High-Density",
      "Urban Low-Density",
      "Suburban High-Density",
      "Suburban Low-Density",
      "Exurban"
    ))
  )

# Confirm classification
table(denver_density$landscape)
summary(denver_density$pop_density)

# ============================================================
# STEP 5 — Create plot versions (WGS84 for ggplot mapping)
# ============================================================
denver_density_plot <- denver_density %>% st_transform(4326)
denver_msa_plot     <- denver_msa     %>% st_transform(4326)

# ============================================================
# STEP 6 — Landscape classification map
# Walker Ch. 6 — ggplot2 geom_sf()
# ============================================================
landscape_map <- ggplot() +
  geom_sf(data  = denver_density_plot,
          aes(fill = landscape),
          color = "white", lwd = 0.1) +
  geom_sf(data  = denver_msa_plot,
          fill  = NA, color = "black", lwd = 0.8) +
  scale_fill_manual(
    values = c(
      "Urban High-Density"    = "#b2182b",
      "Urban Low-Density"     = "#ef8a62",
      "Suburban High-Density" = "#9ecae1",
      "Suburban Low-Density"  = "#d1e5f0",
      "Exurban"               = "#2166ac"
    ),
    name = "Hanberry Landscape\n(Table 4 Worldpop)"
  ) +
  labs(
    title    = "Denver Metro Area: Urban–Suburban–Exurban\nLandscape (2020)",
    subtitle = "Classified using Hanberry (2022) Worldpop thresholds",
    caption  = "Source: 2020 Decennial Census · Hanberry (2022) Table 4"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 13,
                                   margin = margin(b = 4)),
    plot.subtitle   = element_text(size = 10, color = "grey40",
                                   margin = margin(b = 8)),
    plot.caption    = element_text(size = 8,  color = "grey50"),
    legend.position = "right",
    legend.title    = element_text(size = 9,  face = "bold"),
    legend.text     = element_text(size = 8),
    plot.margin     = margin(10, 10, 10, 10)
  )

landscape_map
# ============================================================
# STEP 7 — Download ACS demographic variables
# Walker Ch. 2.1.2 — get_acs(), survey = "acs5"
# Walker Ch. 3 — variable codes explained p.27-34
# ============================================================
denver_acs <- get_acs(
  geography = "tract",
  variables = c(
    total_pop    = "B02001_001",  # race universe
    white        = "B02001_002",  # White alone
    black        = "B02001_003",  # Black alone
    hispanic     = "B03003_003",  # Hispanic/Latino (ethnicity)
    med_income   = "B19013_001",  # median HH income
    below_pov    = "B17001_002",  # below poverty line
    pov_universe = "B17001_001",  # poverty universe
    bach         = "B15003_022",  # bachelor's degree
    masters      = "B15003_023",  # master's degree
    professional = "B15003_024",  # professional degree
    doctorate    = "B15003_025",  # doctorate degree
    edu_universe = "B15003_001"   # education universe (25+)
  ),
  state  = "CO",
  year   = 2020,
  survey = "acs5"
)

# ============================================================
# STEP 8 — Reshape and join ACS to spatial data
# Walker Ch. 3.2 — pivot_wider() and left_join()
# Walker p.34 — Hispanic is ethnicity not race (separate table)
# ============================================================
denver_acs_wide <- denver_acs %>%
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate)

denver_full <- denver_density %>%
  left_join(denver_acs_wide, by = "GEOID") %>%
  mutate(
    pct_white   = white    / total_pop   * 100,
    pct_black   = black    / total_pop   * 100,
    pct_hisp    = hispanic / total_pop   * 100,
    pct_poverty = below_pov / pov_universe * 100,
    pct_college = (bach + masters + professional + doctorate) /
                   edu_universe * 100
  )

denver_full_plot <- denver_full %>% st_transform(4326)

# ============================================================
# STEP 9 — Summary table by landscape
# Walker Ch. 3.3.1 — group_by() and summarize()
# ============================================================
landscape_summary <- denver_full %>%
  st_drop_geometry() %>%
  group_by(landscape) %>%
  summarize(
    total_pop   = sum(pop_total,   na.rm = TRUE),
    med_income  = median(med_income, na.rm = TRUE),
    pct_white   = weighted.mean(pct_white,   pop_total,    na.rm = TRUE),
    pct_black   = weighted.mean(pct_black,   pop_total,    na.rm = TRUE),
    pct_hisp    = weighted.mean(pct_hisp,    pop_total,    na.rm = TRUE),
    pct_poverty = weighted.mean(pct_poverty, pov_universe, na.rm = TRUE),
    pct_college = weighted.mean(pct_college, edu_universe, na.rm = TRUE)
  )

print(landscape_summary, width = Inf)

# ============================================================
# STEP 10 — Income bar chart
# Walker Ch. 4.1 — ggplot2 geom_col()
# Walker Ch. 4.2.3 — ggsave()
# ============================================================
income_chart <- ggplot(
  landscape_summary,
  aes(x = landscape, y = med_income, fill = landscape)
) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = dollar(med_income)), vjust = -0.4, size = 3.5) +
  scale_y_continuous(
    labels = dollar_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_fill_manual(values = c(
    "Urban High-Density"    = "#b2182b",
    "Urban Low-Density"     = "#ef8a62",
    "Suburban High-Density" = "#9ecae1",
    "Suburban Low-Density"  = "#d1e5f0",
    "Exurban"               = "#2166ac"
  )) +
  labs(
    title    = "Median Household Income by Landscape — Denver MSA",
    subtitle = "2016–2020 ACS 5-Year Estimates",
    x        = NULL,
    y        = "Median Household Income",
    caption  = "Source: U.S. Census Bureau ACS"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

income_chart

# ============================================================
# STEP 11 — Poverty rate choropleth map
# Walker Ch. 6 — tmap/ggplot2 choropleth mapping
# ============================================================
denver_poverty_plot <- denver_full %>%
  st_transform(4326) %>%
  select(GEOID, pct_poverty, geometry) %>%
  filter(!is.na(pct_poverty)) %>%
  mutate(
    poverty_cat = cut(
      pct_poverty,
      breaks = c(0, 2.745, 5.189, 8.247, 12.802, 77.935),
      labels = c(
        "Very Low (0–2.7%)",
        "Low (2.7–5.2%)",
        "Moderate (5.2–8.2%)",
        "High (8.2–12.8%)",
        "Very High (12.8–77.9%)"
      ),
      include.lowest = TRUE
    )
  )

poverty_map <- ggplot(denver_poverty_plot) +
  geom_sf(aes(fill = poverty_cat), color = "white", lwd = 0.1) +
  geom_sf(data = denver_msa_plot, fill = NA,
          color = "black", lwd = 0.8) +
  scale_fill_manual(
    values = c(
      "Very Low (0–2.7%)"      = "#fee5d9",
      "Low (2.7–5.2%)"         = "#fcae91",
      "Moderate (5.2–8.2%)"    = "#fb6a4a",
      "High (8.2–12.8%)"       = "#cb181d",
      "Very High (12.8–77.9%)" = "#67000d"
    ),
    name     = "% Below Poverty Line",
    na.value = "grey80"
  ) +
  labs(
    title    = "Poverty Rate by Tract —\nDenver MSA (2020 ACS)",
    subtitle = "Quantile classification · 2016–2020 ACS 5-Year Estimates",
    caption  = "Source: U.S. Census Bureau ACS"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 13,
                                   margin = margin(b = 4)),
    plot.subtitle   = element_text(size = 10, color = "grey40",
                                   margin = margin(b = 8)),
    plot.caption    = element_text(size = 8,  color = "grey50"),
    legend.position = "right",
    legend.title    = element_text(size = 9,  face = "bold"),
    legend.text     = element_text(size = 8),
    plot.margin     = margin(10, 10, 10, 10)
  )

poverty_map

# ============================================================
# STEP 12 — Racial composition bar chart
# Walker Ch. 4 — ggplot2 grouped bar chart
# Walker p.34 — Hispanic as separate ethnic category
# ============================================================
race_summary <- denver_full %>%
  st_drop_geometry() %>%
  group_by(landscape) %>%
  summarize(
    White    = weighted.mean(pct_white, pop_total, na.rm = TRUE),
    Black    = weighted.mean(pct_black, pop_total, na.rm = TRUE),
    Hispanic = weighted.mean(pct_hisp,  pop_total, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols      = c(White, Black, Hispanic),
    names_to  = "group",
    values_to = "pct"
  ) %>%
  mutate(group = factor(group, levels = c("White", "Hispanic", "Black")))

race_chart <- ggplot(
  race_summary,
  aes(x = landscape, y = pct, fill = group)
) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    position = position_dodge(width = 0.7),
    vjust    = -0.4,
    size     = 2.8
  ) +
  scale_fill_manual(
    values = c(
      "White"    = "#2166ac",
      "Black"    = "#b2182b",
      "Hispanic" = "#f4a582"
    ),
    name = "Racial group"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(round(x), "%"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Racial Composition by Landscape — Denver MSA",
    subtitle = "Population-weighted averages · 2016–2020 ACS 5-Year Estimates",
    x        = NULL,
    y        = "Share of population (%)",
    caption  = "Source: U.S. Census Bureau ACS"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13,
                                    margin = margin(b = 4)),
    plot.subtitle    = element_text(size = 10, color = "grey40",
                                    margin = margin(b = 8)),
    plot.caption     = element_text(size = 8,  color = "grey50"),
    axis.text.x      = element_text(angle = 20, hjust = 1),
    legend.position  = "right",
    panel.grid.minor = element_blank()
  )

race_chart

# ============================================================
# STEP 13 — Hispanic population share map
# Walker Ch. 6 — choropleth mapping
# ============================================================
denver_hisp_plot <- denver_full %>%
  st_transform(4326) %>%
  select(GEOID, pct_hisp, geometry) %>%
  filter(!is.na(pct_hisp)) %>%
  mutate(
    hisp_cat = cut(
      pct_hisp,
      breaks = c(0, 10, 20, 35, 50, 100),
      labels = c(
        "Very Low (0–10%)",
        "Low (10–20%)",
        "Moderate (20–35%)",
        "High (35–50%)",
        "Very High (50–100%)"
      ),
      include.lowest = TRUE
    )
  )

hisp_map <- ggplot(denver_hisp_plot) +
  geom_sf(aes(fill = hisp_cat), color = "white", lwd = 0.1) +
  geom_sf(data = denver_msa_plot, fill = NA,
          color = "black", lwd = 0.8) +
  scale_fill_manual(
    values = c(
      "Very Low (0–10%)"    = "#f7f4f9",
      "Low (10–20%)"        = "#d4b9da",
      "Moderate (20–35%)"   = "#c994c7",
      "High (35–50%)"       = "#df65b0",
      "Very High (50–100%)" = "#91003f"
    ),
    name     = "% Hispanic",
    na.value = "grey80"
  ) +
  labs(
    title    = "Hispanic Population Share by Tract —\nDenver MSA",
    subtitle = "2016–2020 ACS 5-Year Estimates",
    caption  = "Source: U.S. Census Bureau ACS"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 13,
                                   margin = margin(b = 4)),
    plot.subtitle   = element_text(size = 10, color = "grey40",
                                   margin = margin(b = 8)),
    plot.caption    = element_text(size = 8,  color = "grey50"),
    legend.position = "right",
    legend.title    = element_text(size = 9,  face = "bold"),
    legend.text     = element_text(size = 8),
    plot.margin     = margin(10, 10, 10, 10)
  )

hisp_map

# ============================================================
# STEP 14 — Population pyramids
# Walker Ch. 4.5 p.79-82 — population pyramids
# B01001 = Sex by Age table
# Males flipped negative per Walker p.81
# ============================================================
male_vars_list   <- paste0("B01001_0", str_pad(3:25,  2, pad = "0"))
female_vars_list <- paste0("B01001_0", str_pad(27:49, 2, pad = "0"))

age_sex <- get_acs(
  geography = "tract",
  variables = c(male_vars_list, female_vars_list),
  state     = "CO",
  year      = 2020,
  survey    = "acs5"
)

male_vars   <- paste0("B01001_0", str_pad(3:25,  2, pad = "0"))
female_vars <- paste0("B01001_0", str_pad(27:49, 2, pad = "0"))

age_labels <- c(
  "Under 5", "5–9", "10–14", "15–17", "18–19",
  "20", "21", "22–24", "25–29", "30–34",
  "35–39", "40–44", "45–49", "50–54", "55–59",
  "60–61", "62–64", "65–66", "67–69", "70–74",
  "75–79", "80–84", "85+"
)

urban_ids <- denver_density %>%
  filter(str_detect(landscape, "Urban")) %>%
  pull(GEOID)

suburban_ids <- denver_density %>%
  filter(str_detect(landscape, "Suburban")) %>%
  pull(GEOID)

exurban_ids <- denver_density %>%
  filter(landscape == "Exurban") %>%
  pull(GEOID)

make_pyramid_data <- function(geoids) {
  age_sex %>%
    filter(GEOID %in% geoids) %>%
    mutate(sex = case_when(
      variable %in% male_vars   ~ "Male",
      variable %in% female_vars ~ "Female",
      TRUE                      ~ NA_character_
    )) %>%
    filter(!is.na(sex)) %>%
    group_by(variable, sex) %>%
    summarize(pop = sum(estimate, na.rm = TRUE), .groups = "drop") %>%
    group_by(sex) %>%
    mutate(age_group = age_labels) %>%
    ungroup() %>%
    mutate(
      age_group = factor(age_group, levels = age_labels),
      pop       = if_else(sex == "Male", -pop, pop)
    )
}

make_pyramid_plot <- function(data, title_text) {
  max_val <- max(abs(data$pop), na.rm = TRUE)
  ggplot(data, aes(x = pop, y = age_group, fill = sex)) +
    geom_col(width = 0.85, alpha = 0.85) +
    scale_x_continuous(
      labels = ~ scales::number(abs(.x),
                                scale    = 1e-3,
                                suffix   = "k",
                                accuracy = 1),
      limits = c(-max_val, max_val),
      breaks = pretty(c(-max_val, max_val), n = 6)
    ) +
    scale_fill_manual(
      values = c(Female = "#b2182b", Male = "#2166ac")
    ) +
    geom_vline(xintercept = 0, color = "white", lwd = 0.8) +
    labs(
      title    = title_text,
      subtitle = "Denver MSA · 2016–2020 ACS",
      x        = "Population",
      y        = "Age Group",
      fill     = "",
      caption  = "Source: U.S. Census Bureau ACS B01001"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", size = 13,
                                      margin = margin(b = 4)),
      plot.subtitle    = element_text(size = 10, color = "grey40",
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = 8,  color = "grey50"),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
}

urban_pyramid_data    <- make_pyramid_data(urban_ids)
suburban_pyramid_data <- make_pyramid_data(suburban_ids)
exurban_pyramid_data  <- make_pyramid_data(exurban_ids)

pyramid_urban <- make_pyramid_plot(
  urban_pyramid_data,
  "Population Pyramid — Urban Tracts\nDenver MSA"
)

pyramid_suburban <- make_pyramid_plot(
  suburban_pyramid_data,
  "Population Pyramid — Suburban Tracts\nDenver MSA"
)

pyramid_exurban <- make_pyramid_plot(
  exurban_pyramid_data,
  "Population Pyramid — Exurban Tracts\nDenver MSA"
)

pyramid_urban
pyramid_suburban
pyramid_exurban



# ============================================================
# STEP 15 — Confirm all output files saved
# ============================================================
list.files(pattern = "\\.png$")
# (keep all your existing code here, just no ggsave lines needed)
```

## Overview

This study examines demographic change and spatial inequality across the Denver–Aurora–Lakewood Metropolitan Statistical Area. With a population of nearly 3 million residents, Denver is one of the fastest-growing metropolitan regions in the United States and represents more than half of Colorado's population. As the region has expanded, demographic and socioeconomic differences have emerged across urban, suburban, and exurban landscapes.
Using data from the 2020 Decennial Census and the 2016–2020 American Community Survey (ACS), census tracts are classified into five landscape types: urban high, urban low, suburban high, suburban low, and exurban, using population density thresholds adapted from Hanberry (2022). This framework allows demographic patterns to be examined across a continuous urban–rural gradient rather than relying solely on administrative boundaries. The analysis explores how income, poverty, racial composition, and age structure vary across these landscape types and what these patterns reveal about metropolitan growth in Denver.


## Landscape Classification

Urban high-density tracts concentrate in the city core while suburban 
and exurban tracts extend toward the periphery. Although exurban areas occupy the largest share of land, most residents live within the urban core, illustrating the uneven relationship between population concentration and land area.

```{r}
landscape_map
```

## Income

Median household income increases from urban to exurban landscapes, suggesting that higher-income households are concentrated in lower-density parts of the metropolitan region

```{r}
income_chart
```

## Poverty
Poverty is most concentrated in the urban core, particularly central Denver. As you move outward from the core, poverty rates generally decline, with much of the suburban ring showing lower poverty rates. However, the pattern is not uniform; there are still scattered pockets of higher poverty in outer and peripheral areas, including parts of the southwest metro and more exurban tracts. Overall, Denver shows a strong urban concentration of poverty, combined with a more uneven and dispersed poverty pattern outside the core, rather than a smooth decline.

```{r}
poverty_map
```

## Racial Composition

Hispanic residents are concentrated in urban and inner suburban areas, where they represent the largest minority populations across the metropolitan region. In contrast, suburban and exurban landscapes remain majority White, reinforcing a persistent spatial gradient in racial composition across metropolitan space.

```{r}
race_chart
hisp_map
```

## Age Structure

Urban tracts are characterized by a younger population structure, suburban tracts by a concentration of family households and working-age adults, and exurban tracts by an older population profile.

```{r}
pyramid_urban
pyramid_suburban
pyramid_exurban
```

## What the Data Reveal

Overall, the findings highlight a consistent spatial gradient in which socioeconomic status increases with decreasing density. However, this relationship is mediated by housing tenure, life stage, and racial composition, suggesting that metropolitan growth in Denver is unevenly distributed across both space and social groups.
